---
title: "Churn Statistical Analysis"
author: "Ruben Cabrera"
date: "2026-04"
output: 
  github_document: default
  html_document: default
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load Libraries

```{r libraries}
required_pkgs <- c("readxl", "dplyr", "ggplot2", "caret", "randomForest", "broom")
new_pkgs <- required_pkgs[!required_pkgs %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

library(readxl)
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)
library(broom)
```

## Load & Merge Data

Both files share the `Phone` and `Area Code` columns, which are used as join keys.

```{r load-merge}
calls    <- read_xls("CallsData.xls")
contract <- read.csv("ContractData.csv", stringsAsFactors = TRUE)

# Standardise key column names before joining
# contract already reads "Area Code" as Area.Code; align calls to match
calls <- calls %>% rename(Area.Code = `Area Code`)

churn <- inner_join(contract, calls, by = c("Phone", "Area.Code"))

# Sanitise all column names (replace spaces/special chars with dots)
# so that formula-based models (glm, randomForest) work without issues
names(churn) <- make.names(names(churn), unique = TRUE)

# Convert binary columns to factors for cleaner plots
churn <- churn %>%
  mutate(
    Churn       = factor(Churn,       labels = c("No Churn", "Churn")),
    Int.l.Plan  = factor(Int.l.Plan,  labels = c("No Intl Plan", "Intl Plan")),
    VMail.Plan  = factor(VMail.Plan,  labels = c("No VMail", "VMail"))
  )

glimpse(churn)
```

## Plot 1 — Churn Distribution

```{r plot-churn-dist}
ggplot(churn, aes(x = Churn, fill = Churn)) +
  geom_bar(width = 0.5, show.legend = FALSE) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Customer Churn Distribution",
    x     = NULL,
    y     = "Count"
  ) +
  theme_minimal()
```

## Plot 2 — Day Minutes by Churn

```{r plot-day-mins}
ggplot(churn, aes(x = Churn, y = Day.Mins, fill = Churn)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Day Minutes by Churn Status",
    x     = NULL,
    y     = "Day Minutes"
  ) +
  theme_minimal()
```

## Plot 3 — Customer Service Calls by Churn

```{r plot-custserv}
ggplot(churn, aes(x = factor(CustServ.Calls), fill = Churn)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Churn Rate by Number of Customer Service Calls",
    x     = "Customer Service Calls",
    y     = "Proportion",
    fill  = NULL
  ) +
  theme_minimal()
```

## Plot 4 — International Plan vs Churn

```{r plot-intl-plan}
ggplot(churn, aes(x = Int.l.Plan, fill = Churn)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Churn Rate by International Plan",
    x     = NULL,
    y     = "Proportion",
    fill  = NULL
  ) +
  theme_minimal()
```

## Plot 5 — Total Charges Distribution by Churn

```{r plot-total-charges}
churn <- churn %>%
  mutate(Total.Charge = Day.Charge + Eve.Charge + Night.Charge + Intl.Charge)

ggplot(churn, aes(x = Total.Charge, fill = Churn)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Total Charge Distribution by Churn Status",
    x     = "Total Charge ($)",
    y     = "Density",
    fill  = NULL
  ) +
  theme_minimal()
```

## Plot 6 — Account Length vs Total Charge

```{r plot-scatter}
ggplot(churn, aes(x = Account.Length, y = Total.Charge, colour = Churn)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(
    title  = "Account Length vs Total Charge",
    x      = "Account Length (months)",
    y      = "Total Charge ($)",
    colour = NULL
  ) +
  theme_minimal()
```

---

## Predictive Modelling: Estimating Churn

The target variable `Churn` is binary (No Churn / Churn), so we need a **classification** model.
Two well-suited candidates for this problem are:

### Option 1 — Logistic Regression

Logistic Regression models the **probability** that a customer churns as a function of the
predictors, using the logistic (sigmoid) function to keep outputs in \([0, 1]\):

\[
P(\text{Churn} = 1 \mid \mathbf{x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \cdots + \beta_p x_p)}}
\]

**Advantages:**

- Highly interpretable — coefficients express the log-odds change per unit predictor increase.
- Fast to train, works well even with relatively small datasets.
- Provides calibrated probability estimates out of the box.

**Limitations:**

- Assumes a linear relationship between predictors and the log-odds of the outcome.
- Sensitive to highly correlated predictors (multicollinearity).

### Option 2 — Random Forest

Random Forest builds an **ensemble of decision trees**, each trained on a random bootstrap
sample of the data and a random subset of features. The final prediction is the majority vote
across all trees.

**Advantages:**

- Captures non-linear relationships and interaction effects automatically.
- Robust to outliers and does not require feature scaling.
- Provides variable importance scores, helping identify the key drivers of churn.

**Limitations:**

- Less interpretable than Logistic Regression ("black-box" nature).
- More computationally expensive to train and tune.

---

### Model Preparation

We proceed with **Logistic Regression** as the primary model given its interpretability,
and use **Random Forest** as a complementary model to capture non-linear patterns.
Both are evaluated on a held-out test set (30 % of the data).

```{r model-prep}
library(caret)

# Re-encode Churn as a clean binary factor (required by caret)
churn_model <- churn %>%
  mutate(Churn = factor(Churn, levels = c("No Churn", "Churn")))

# Drop identifier columns not useful for prediction
churn_model <- churn_model %>%
  select(-Phone, -State, -Area.Code)

# Train / test split (70 / 30)
set.seed(42)
train_idx <- createDataPartition(churn_model$Churn, p = 0.70, list = FALSE)
train_set <- churn_model[ train_idx, ]
test_set  <- churn_model[-train_idx, ]

cat("Train rows:", nrow(train_set), "| Test rows:", nrow(test_set), "\n")
```

---

## Model 1 — Logistic Regression

```{r logistic-regression}
log_model <- glm(Churn ~ ., data = train_set, family = binomial)

summary(log_model)
```

```{r logistic-eval}
log_probs <- predict(log_model, newdata = test_set, type = "response")
log_preds <- factor(ifelse(log_probs >= 0.5, "Churn", "No Churn"),
                    levels = c("No Churn", "Churn"))

cm_log <- confusionMatrix(log_preds, test_set$Churn, positive = "Churn")
print(cm_log)
```

```{r logistic-importance}
# Visualise significant predictors via odds ratios
log_coefs <- broom::tidy(log_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(estimate))

ggplot(log_coefs, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3, colour = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  coord_flip() +
  labs(
    title = "Logistic Regression — Odds Ratios with 95% CI",
    x     = NULL,
    y     = "Odds Ratio"
  ) +
  theme_minimal()
```

---

## Model 2 — Random Forest

```{r random-forest}
library(randomForest)

set.seed(42)
rf_model <- randomForest(Churn ~ ., data = train_set, ntree = 300, importance = TRUE)

print(rf_model)
```

```{r rf-eval}
rf_preds <- predict(rf_model, newdata = test_set)

cm_rf <- confusionMatrix(rf_preds, test_set$Churn, positive = "Churn")
print(cm_rf)
```

```{r rf-importance}
# Variable importance plot
varImpPlot(rf_model, main = "Random Forest — Variable Importance")
```

---

## Model Comparison

```{r model-comparison}
results <- data.frame(
  Model     = c("Logistic Regression", "Random Forest"),
  Accuracy  = c(cm_log$overall["Accuracy"],  cm_rf$overall["Accuracy"]),
  Precision = c(cm_log$byClass["Precision"], cm_rf$byClass["Precision"]),
  Recall    = c(cm_log$byClass["Recall"],    cm_rf$byClass["Recall"]),
  F1        = c(cm_log$byClass["F1"],        cm_rf$byClass["F1"])
)

knitr::kable(results, digits = 4, caption = "Test-set performance comparison")
```

---

## Conclusions

### Overall Performance

Random Forest clearly outperforms Logistic Regression across every metric on the held-out
test set (30 % of 3,333 observations):

| Metric    | Logistic Regression | Random Forest |
|-----------|--------------------:|-------------:|
| Accuracy  | 86.19 %             | **98.00 %**  |
| Precision | 55.56 %             | **100.00 %** |
| Recall    | 20.83 %             | **86.11 %**  |
| F1        | 0.303               | **0.925**    |

### Logistic Regression

Although it achieves a seemingly acceptable accuracy of **86.2 %**, this figure is misleading
given the class imbalance in the dataset (~14 % churners). The model struggles to identify
actual churners: its **Recall of only 20.8 %** means it misses roughly 4 out of every 5
customers who will churn — a critical failure for a retention use case where false negatives
are costly. Its low F1 score (0.30) confirms that it is not a reliable churn detector.

### Random Forest

Random Forest achieves **98 % accuracy** with a **Precision of 100 %** (no false alarms) and
a **Recall of 86.1 %** (detects the vast majority of true churners), yielding an F1 of **0.93**.
This makes it the clearly preferred model for deployment in a churn-prevention campaign.

### Key Drivers of Churn (Random Forest Variable Importance)

The top predictors ranked by Mean Decrease in Gini impurity are:

1. **Total Charge** (136.17) — the single strongest predictor; heavy spenders are more likely to leave.
2. **Customer Service Calls** (68.97) — repeated contacts signal dissatisfaction and strongly anticipate churn.
3. **Day Charge / Day Mins** (~46) — high daytime usage costs drive customers away.
4. **International Plan** (40.08) — customers on the international plan churn at a disproportionately high rate.
5. **VMail Messages** (27.33) — lower voicemail engagement is associated with churn.

### Recommendations

- **Prioritise high-billing customers** for proactive retention offers before they reach the churn threshold.
- **Flag customers with ≥ 3 service calls** for immediate intervention; dissatisfaction escalates quickly.
- **Review the International Plan pricing** — its strong association with churn suggests it may not deliver perceived value.
- For production use, consider **threshold tuning** on the Random Forest probability output to further optimise
  the Recall/Precision trade-off according to the cost of false negatives vs. false positives.
